The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts

### 2020-05-21 write-up/scifile.tex
Update scifile.tex

\section*{Introduction}

The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.  As of May 21st, 2020, over 5 million cases have been confirmed worldwide with 1.5 million confirmed cases and over 90 thousand COVID-19 deaths across the United States. This pandemic
%has ripped through our communities and
has become the focal point of every day life; yet the data landscape for understanding COVID-19 remains limited.  Public databases maintained by Johns Hopkins University~\cite{JHU_Lancet} and the New York Times~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists have attempted to use this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\cite{Song2020,Ray2020,2020.IHME}.

@@

\section*{Analysis of imperfect case-count data}
\label{section:casecount}

@@

In the current context, the true status is not observed. Instead, individuals are tested for COVID-19 status. Tests are imperfect.  COVID-19 testing is no exception. The standard assumption is that measurement error leads to parameter attenuation.  When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs~\cite{Beesley2020,Beesley2019,Smeden2019}. Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.  Suppose disease prevalence is estimated by $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling. We wish to express the error $\tilde  y_n - \bar Y$ in statistical terms. In the supplementary materials, we show that the error now can be expressed as
\begin{equation}
\label{eq:error}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
\end{equation}

The above selection mechanisms are random and independent of the outcome of interest. When this is not the case, selection effects may cause bias in the above estimates. To better understand this issue, \cite{Meng2018} provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
$$
\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
$$
The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}. The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.  Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.  The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.  The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.  Meng identified this as the fundamental issue that can lead to paradoxes in the analysis of big data.

@@

The remainder of this paper aims to build upon these fundamental insights by extending the decomposition in two directions: accounting for  measurement error and the temporal nature of the pandemic.  We then discuss effective sampling methods and their importance in the current pandemic.

@@

Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.  Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.  We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
$$
\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
$$
where $Z = 1-2Y$; see Appendix~\ref{app:imperfect} for the derivation. The first term in the large brackets represents the perfect testing regime; to this end, we refer to $\rho_{I,Y}$ as the \emph{true data quality}.  The second term represents the interaction between imperfect testing and selection bias. The variable $PZ = 1[Y=0,P=1] - 1[Y=1, P=1]$ is non-zero only when $P=1$ (i.e., the outcome is incorrectly reported); to this end, we refer to $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement error.  In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.  Finally, the third term represents the bias due to imperfect testing.

We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement error and selection bias.  To do this, we define the sampling rates differential.  Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ be the sampling rates.  Then $\Delta = f_1 - f_0$ is the sampling rate differential.  Using these terms, we can re-express the first two terms as
$$
\rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} =
\rho_{I,Y} \times \sigma_Y \left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
$$
The final term in brackets is the \emph{measurement error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence. Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the appendix, which implies the measurement error adjustment either shrinks the data quality measure toward zero or reverses its sign.

While prior investigations have noted the interaction between measurement error and selection bias~\cite{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.  The above statistical decomposition clarifies the importance of this quantity~$f$.  In particular, note that the statistical error also includes a bias term due to measurement error and this term increases as the sampled fraction $f$ increases. Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.  This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.

Consider again the current COVID-19 pandemic. For now, we continue to assume the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.  In Section~\ref{section:est_dq}, we discuss recent research suggesting a false negative rate around 17.2\% and a false positive rate around 0.05\%. Under these rates and the current US prevalence rates, the ratio of the MSE to MSE under no measurement error is 0.436; if we switch false negative rates down to 5\% and increase the false positive rate to 5\% then the relative MSE is 2.89.  This is merely to demonstrate that in some cases we see a huge increase in MSE and in other settings we have a huge decrease in MSE.  What drives this is the false positive and negative rate interaction with prevalence and sampling rates.  Therefore, whether we are better or worse off with respect to the MSE is a very difficult question to answer.

The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may suggest the alternative estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling. We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms. In the appendix, we show that the error now can be expressed as
\begin{equation}
\label{eq:error}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
\end{equation}
The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement error.  The second term is the interaction between selection bias and measurement error.  Figure~2 presents the measurement error data quality adjustment as a function of the relative frequency and odds ratio.  We see that the adjustment can be both positive and negative as well as a range of magnitudes.

>>>>

\paragraph*{Introduction}

The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.  As of May 21st, 2020, over 5 million cases have been confirmed worldwide with 1.5 million confirmed cases and over 90 thousand COVID-19 deaths across the United States. This pandemic
has become the focal point of every day life; yet the data landscape for understanding COVID-19 remains limited.  Public databases maintained by Johns Hopkins University~\cite{JHU_Lancet} and the New York Times~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists have attempted to use this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\cite{Song2020,Ray2020,2020.IHME}.

@@

\paragraph*{Analysis of imperfect case-count data}

@@

In the current context, the true status is not observed. Instead, individuals are tested for COVID-19 status. Tests are imperfect.  COVID-19 testing is no exception. Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.  We suppose disease prevalence is estimated by $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling.

To better understand this issue, we provide the following intuitive and powerful statistical decomposition of the error between $\tilde y_n$ and the true proportion $\bar Y$
\begin{equation}
\label{eq:error}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M},
\end{equation}
where $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ are sampling rates for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.  This extends prior work by \cite{Meng2018} to account for imperfect testing. The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.  The final term in brackets is the \emph{measurement error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence. Figure~2 presents the measurement error data quality adjustment as a function of the relative frequency and odds ratio.  We see that the adjustment can be both positive and negative as well as a range of magnitudes.

The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.  Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.  The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.  The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$. We can see here  selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs~\cite{Beesley2020,Beesley2019,Smeden2019}.

@@

@@
